Optimally Efficient Collaborative Text Editing

### 2012-12-17 paper-collabed.tex
completed major revision

The data type $M$ allows the following operations:
\begin{itemize}
\item $M.treeextend(v/m:p,\alpha)$ - lookup the node with nodeid v/n:p and extend it by adding the character $\alpha$ at position $p$, update the lists $M.S$, $M.R$, and $M.E$ as well.
\item $M.treehide(v/m:p,q)$ - set the visibility to false for the element at position $q$ of the node with id $v/m:p$
\item $M.treeinsert(v/m:p,q,u/n,\alpha)$ - create a new node with id $u/n:1$ containing the single visible character $/alpha$ and insert that new node at offset $q$ in the node with id $v/m:p$.
\item $M.converttotreeop(\sigma)$- returns a tree-edit operation $\tau$ on $M.T$ that corresponds to the string edit operation $\sigma$ on $M.S$
\item $M.converttostringop(\tau,X)$ - return the string edit operation (and insertion or deletion) for the string $M.X$ which correspond to applying the tree edit operation $\tau$ on $M.T$
\end{itemize}
The three tree operations also a triple of string operations that can be applied to
the views $M.S$, $M.R$, and $M.E$ so they will agree with the modified $M.T$.
We will provide algorithms for all five of these operations and prove that their complexity is $O(\log(N))$.
\subsection{implementation of characters and markers}
We assume all visible characters come from an alphabet $\Sigma$
and we expand that alphabet to $\Sigma^*$ by including all start and endmarkers.
The elements in the lists and the tree will be represented by a tuple that contains links to the nodes and also links to an indexing object $M.I$ which will allow one to rapidly calculate the position of the element in the lists (i.e. the number of elements of type $S$, $R$, or $E$ to the left of the element) and also allows us to rapidly find the $k$th element of each of the three lists.

@@

@@

@@

@@

@@

\subsection{applying $treeinsert(v/m:p,q,u/n,\alpha)$ to $M.T$}
This is the trickiest tree edit operation and requires several cases..

\subsubsection{converting $treeinsert(v/m:p,q,u/n,\alpha)$ to a string op}
Again, we lookup the node $n$ with id $v/m:p$ and the element $e$ at position $q+1$ if $q<p$ or the end position $e$ if $q=p$.

\subsection{Main Theorem}

@@

\section{The unoptimized model for edit trees with a edit-string view}
We are really interested in collaborative editing of strings, not trees, so the
question arises as to how we can convert string edit operations of the user
into tree edit operations on the user's model, and how we can convert tree edit
operations from the user's peers into string edit operations on the user's string.

As a first step, observe that there is a one-to-one correspondence between
edit-strings as defined above and edit-trees, thus our distributed algorithm
for edit-trees provides a distributed algorithm for edit-strings where we
assume that the clients $C_i$ each maintain a edit-string $S_i$ and perform valid
edit-string operations on $S_i$. These operations can be transformed directly
into valid tree operations which can be broadcast to the peers above.

This raises the question of which string edit operations are valid operations,
in the sense that they correspond to the valid tree-edit operations described
in the previous section.  Clearly any non-marker node can be "deleted" by setting
its visibility property to "hidden."  The insertion operations are a little
trickier, but its not hard to show that the following operations are precisely
those which correspond to valid-tree operations.

We describe the valid edit-string insertion operations as insertions
between two symbols $d_i$ and $d_{i+1}$ in the edit-string, where the $d_i$
is one of the following:
\begin{itemize}
\item  a character, $c_i$, either visible or hidden,
\item  a start marker $<^v_m$ or
\item an end marker $>^v_m$
\end{itemize}
Based on this categorizations, there are nine possibilites for inserting a character between two adjacent elements in an edit string. In each case we could either enter a single character $\alpha$ or a node $<^u_n \alpha >^u_n$ containing that character. The next lemma considers all of those possibilities and specifies which ones correspond to legal operations on an edit tree.

Let $u$ denote the user performing
the operation and let $n$ be the next available block id for user $u$, so if
the user inserts a new block containing the character $\alpha$, it will be inserted
as $<^u_n \alpha >^u_n$.  Let $c_i$ and $c_j$ denote non-marker characters in
a block $v/m$ 
which could be either visible or hidden or one of each and let
 $v$ and $w$ be users different from $u$ and each other.

\begin{lemma} Notation as above. There are 9 cases for inserting a character or a node containing a character into an edit string between two elements $c_i$ and $c_{i+1}$ in an edit string within the scope of a node $v/m$. The following cases specify which possibilites correspond to valid operations on the edit tree and which do not. If a case is not listed, then it is not valid. 
\begin{eqnarray}
c_i\;\;\;\;\; c_{i+1} &\rightarrow& c_i <^u_n \alpha >^u_n c_{i+1} \label{eq:ins1}\\
c_i\;\; >^w_r &\rightarrow& \left\{ 
  \begin{array}{l l}
      c_i \alpha >^w_r &  {\rm if}\; u=w  \label{eq:ins2a}\\
      c_i <^u_n \alpha >^u_n >^w_r &  {\rm if} \; u \ne w 
  \end{array}  \right . \label{eq:ins2b}\\
c_i \;\;<^w_r &\rightarrow& \left\{ 
\begin{array}{l l}
c_i <^u_n \alpha >^u_n <^w_r & {\rm if}\; u<w  \label{eq:ins3a}\\
{\rm INVALID} & {\rm if} u \ge w \label{eq:ins3b}\\
\end{array}
\right . \label{eq:ins3}\\
>^v_m \;\;c_{i+1} &\rightarrow& 
 \left\{ 
  \begin{array}{l l}
    >^v_m  <^u_n \alpha >^u_n c_{i+1} & {\rm if} \; v<u \label{eq:ins4a}\\
   {\rm INVALID} & if v \ge u \label{eq:ins4b}\\
  \end{array} \right .\label{eq:ins4}\\
>^v_m \;\;>^w_r &\rightarrow& 
  \left\{
    \begin{array}{l l}
      >^v_m \alpha >^w_r &{\rm if}\; u=w \label{eq:ins5a} \\
      >^v_m <^u_n \alpha >^u_n >^w_r & {\rm if} \; (v<u)  \&  (v \ne w)  \label{eq:ins5b} \\
     {\rm INVALID} & {\rm (v > u) \& (v \ne w)} \label{eq:ins5c}
    \end{array}\right . \label{eq:ins5} \\
>^v_m \;\;<^w_r &\rightarrow & 
\left \{
\begin{array}{l l}
 >^v_m <^u_n \alpha >^u_n <^w_r & {\rm if} \; v<u<w  \label{eq:ins6a}\\
{\rm INVALID} & {\rm otherwise}  \label{eq:ins6b}
\end{array} \right . \label{eq:ins6}\\
<^v_m \;\;c_0 &\rightarrow& <^v_m  <^u_n \alpha >^u_n c_0\label{eq:ins7}\\
<^v_m \;\;>^w_r &\rightarrow& 
\left \{
\begin{array}{l l}
<^v_m <^u_n \alpha >^u_n >^w_r & {\rm if} \; v=w=0, r=m=0 \label{eq:ins8a}\\
{\rm INVALID} & {\rm otherwise} \label{eq:ins8b} \\
\end{array} \right . \label{eq:ins8}\\
<^v_m \;\;<^w_r &\rightarrow&
\left \{
\begin{array}{l l}
 <^v_m <^u_n \alpha >^u_n <^w_r & {\rm if} \; u<w  \label{eq:ins9a}\\
{\rm INVALID} & {\rm otherwise}  \label{eq:ins9b}\\
\end{array} \right . \label{eq:ins9}
\end{eqnarray}
\end{lemma}

\begin{proof}
Case (1): Note that user $u$ can always insert a new node between any two non-marker
characters (Eq. \ref{eq:ins1}), using operation $treeinsert(v/m:p,q,u/n,\alpha)$ but
since the position is not at the end of the node a straight insertion of $\alpha$ is not allowed.

Cases (2) and (5): User $u$ can always insert a string
directly into the end of a node that it owns via $treeextend(v/m:p,\alpha)$
 (Eqs. \ref{eq:ins2a}, \ref{eq:ins5a}) 
but the user can also insert a node at that position if they want $treeinsert(u/n:p,p,u/n,\alpha)$ if no node has been inserted there yet
 (Eq. \ref{eq:ins2b}). If another node has already been inserted at the
last position by user $v$ then user $u$ can only insert a node there if $v<u$.

Case (3):
A node $u/n$ can be inserted as the first node at an insertion point if the current
first node in the insertion point $w/m$ satisfies $u<w$.

Case (4):
A node $u/n$ can be placed at the end of an insertion point where the last node in the insertion is $v/m$ if and only if $v<u$.

Case (6):
A node $u/n$ can be inserted between two nodes $v/m$ and $w/r$ in an insertion point if and only if $v<u<w$.

Case (7): a node $u/n$ can always be placed at position 0 of a node $v/m$ provided nothing has yet been inserted there.

Case (8): 
The only case where a start symbol is immediately followed by an end symbol
if the root case $<^0_0>^0_0$ and in this case, user 0 is the root and does not insert any elements itself, so the only thing that can be inserted is a node, i.e. not a character (Eqs. \ref{eq:ins8a}, \ref{eq:ins8b} ).

Case (9): a node $u/n$ can be inserted at the position 0 of a node $v/m$ which already has some nodes inserted there if and only if the first node inserted a position 0, $w/r$ satisfies $u<w$.

\end{proof}

For now, we observe that if a edit-string view only allows the insertions
as shown above for user $u$, then each such insertion corresponds to a valid
edit-tree operation and hence the collaborative editing protocol in the previous
section for edit-trees can be naturally extended to edit-strings.  The converse
is also true, every valid edit-tree operation corresponds to one of the 9 valid cases above.

The following
lemma states the convergence property for this distributed algorithm.

\begin{lemma}
Let $G$ denote a set of clients $\{C_1,\ldots,C_n\}$
that are running the collaborative
editing protocol for their local copies of edit-strings $S_i$
starting with empty strings.
Suppose they perform some number of valid
edit operations and then stop editing and process all operations in their queues.
Then, 
we will have $S_1=S_2=\ldots=S_n$.
\end{lemma}

\section{The unoptimized model for edit trees with a standard-string view}

This model is only slightly more complicated than the previous case of
collaborative editing of edit-strings.  In this case, each client maintains three
views of the editing process: the standard view of the string currently being edits,
the "tracking changes" view which shows both the visible and "deleted" characters (with a different font of the deleted characters), and the edit-string view which
has characters (both visible and hidden with appropriate attributes for each type)
and markers (both start and end).

Our first observation is that each character in the standard view corresponds
to a uniquely defined visible character in the tracking-changes view, and
each character in the tracking-changes view corresponds to a non-marker character in the edit-string view. So we are really maintaining one list and two nested sublists of that list.

There are two operations we allow on the standard view:
\begin{itemize}
\item ${\tt insert}(p,\alpha)$ - insert the character $\alpha$ 
at offset $p$ in the string
\item ${\tt remove}(p,\alpha)$ - remove the character $\alpha$ at offest $p$
which has the effect of changing the attribute of that character from visible to hidden.
\end{itemize}

If $T$ is a edit-tree and $S = \Gamma(T)$ denotes the standard string view
of that tree, then we want to show that for each string operation $\sigma$
on $S$ there is a corresponding valid operation $\tau$
on the edit-tree, such that 
\[
\Gamma(\tau(T)) = \sigma(\Gamma(T))
\]
This will allow us to convert string operations into edit-tree operations which
can then be broadcast to all users as in the previous section. There can be more than one possible $\tau$ operation that satisfies this property (and often there are many), but all we need is the existence of at least one such operation. At this point, we don't worry about the computation complexity of calculating the edit-tree operation corresponding to a string operation, that will be the subject of the next section.

The remove operation is easily transformed into a edit-tree hide operation which 
simply sets the visibility attribute of the corresponding character to "hidden''.

The insert operation is a bit trickier mainly because, as we have
seen in the example at the beginning of this paper,
a single insert position in
the standard view could correspond to many different insert positions in the 
edit-tree, not all of them legal.

We now describe how to find one such $\tau$ and in the next section we present a data type in which that $\tau$ can always be constructed in time $O(\log(N))$.

There are four basic insert operations on the string $S=\Gamma(T)$ in the standard view to consider.
\begin{itemize}
\item inserting into an empty string
\item inserting at the beginning of the string
\item inserting at the end of the string
\item inserting somewhere in the middle of the string 
\end{itemize}
The first case is easily handled as it results in the following operation on the 
edit-string $\gamma$:
\[
<^0_0\;>^0_0 \;\; \rightarrow \;\;<^0_0 <^u_0 \alpha >^u_0 >^0_0
\]
The second is a little trickier. We need to find the first non-marker character $d_k$
in the full-edit string and observe that the previous character must be a start
marker $<^v_m$ so we can insert the new node $u/n$ containing $\alpha$ between those
two points, where $n$ is the number of nodes owned by $u$ in $T$ and observe
that inserting a new node 
\[
<^u_n \alpha >^u_n
\]
between a start marker and a non-marker character is always valid.

The third case of inserting at the end of the standard view is similar. One
must find the last non-marker character (visible or hidden) $d_k$ in the
edit-string $\gamma$ and observe that $d_{k+1}$ must be an end marker
$>^v_m$ thus we can insert $u/n$ between $d_k$ and $d_{k+1}$ as a valid
edit-string operation.

The last case is when we are inserting between two characters $c_i$ and $c{_i+1}$
in the standard view $S = \Gamma(T)$. Let $d_j$ denote the visible non-marker
character corresponding to $c_i$ in the edit-string $T$ and let
$d_k$ denote the next non-marker character. So $d_j$ and $d_k$ are non-marker
characters, all the symbols between them are marker symbols, $d_j$ is
visible, and $d_k$ could be visible or hidden. There are three subcases:
\begin{itemize} 
\item $d_{j+1}$ is a non-marker character, i.e. $d_j$ and $d_k$ are adjacent. 
In this case, inserting
the node $u/n$ between them is always valid.
\item $d_{j+1}$ is an end marker $>^v_m$, again in this case inserting $u/n$
between $d_j$ and $d_{j+1}$ will be a valid insert operation.
\item $d_{j+1}$ is a start-marker. In this case all of the symbols between
$d_j$ and $d_k$ must be start markers, so $d_{k-1} = <^v_m$ is a start marker
and inserting between that symbol and the non-marker character that follows
is a valid operation.
\end{itemize}
In each of these three cases, we get a edit-string operation $\tau$ that
corresponds to inserting $\alpha$ between $c_i$ and $c_{i+1}$. The arguments
above have proven the following lemma.

\begin{lemma}
Let $T$ be a edit-tree and let $S=\Gamma(T)$ be its standard string view, then
the algorithm described above constructs a valid edit-tree operation $\tau$
on $T$ for each valid edit operation $\sigma$ on the string $S=\Gamma(T)$,
we have
\[
\sigma(\Gamma(T)) = \Gamma(\tau(T))
\]
\end{lemma}

We can now define a
collaborative editing algorithm where the user $u$ presents 
a standard view of a
string $S^u_i$ representing the effect of $i$ editing operations.
The user can then either insert or delete a character from that string
and broadcast the corresponding edit-tree operation to all peers.
Likewise, edit-tree operations that are received from peers can be applied to
the tree results in a modifed string $S^u_{i+1}$. Since the edit operations commute as long as they are independent, the order in which they are applied doesn't matter!
This proves the following theorem.

\begin{theorem}
The full-edit-tree collaborative editing algorithm as described above
is convergent, that is, if all users start with the same tree $T_i=T$ and
all perform a set of editing operations as described above, then after all
editing has stopped and all messages have been received and processed, all clients
will have processed the same number $n$ of edit operations (some local and
some remote) and all will have the same edit tree stored in their local copy
$T_i$.
\end{theorem}

So far we have described a relatively simple collaborative editing algorithm and
proved that it is correct . We have not shown however how
to implement it efficiently. A naive implementation would require time $O(N)$ for
each step, where $N$ is the size of the edit-string. (This would happen
for example if the edit-tree is comb-shaped which would require time $O(N)$
to traverse from the root to the farthest leaf.)

We will show in the 
next section, that this time can be reduced to $O(k\log(N))$ where $k$ is the
size of the operation (i.e. number of characters inserted or deleted) and
$N$ is the size of the edit-string (i.e. the number of characters inserted
and/or deleted by the collaborative users).

\section{Optimizing the model}
% Here we describe the subnode implementation with red-black trees and the initial and
% final subnodes for each node. We prove that the operations are optimally efficient.

We have shown that edit-trees can be used to implement a collaborative editor
which allows the user to immediately apply local edits and which has the convergence
property.  The model as presented in the previous section becomes unacceptably slow ($O(N)$ for each insert and delete) as the size of the tree grows if we use a naive data structure to represent the three lists and the edit tree.

In this section, we show how to optimize this model by creating a more complex representation which allows
the string to tree and tree to string conversions on edit operations to be computed
in time $O(\log(N))$ where $N$ is the size of the edit tree.

In particular, the MSET algorithm will maintain the edit tree $T$ as described above,
but will also maintain an auxiliary data structure $M$ that allows it to rapidly
compute the node position in the edit tree corresponding to an element (i.e. a visible or hidden character or a start or end marker) in any one
of three string views, and vice versa. The three string views are
\begin{itemize}
\item $M.S$: the "standard" list of all visible characters
\item $M.R$: the "tracking changes" list of all visible and hidden characters which shows all of the revision of the document,
\item $M.E$: the "edit-string" list of all edit-string characters which faithfully represents the edit-tree by showing all visible and hidden characters and all markers (both
start markers and end markers)
\end{itemize}
We will show that these three lists can be implemented in such a way as to be able
to calculate the $k$th element of each list (e.g. M.E[k]) in time $O(\log(N))$ where $N$ is the size of the edit-tree. Moreover, given an element $e$ its position in any one of these lists can also be computed in time $O(\log(N))$ using $M.pos(e,L)$ where $L$ is one of $S,R,E$.  In case, $e$ doesn't appear in a list of type $L$ (e.g. a marker character in the $S$ or $R$ list) we can also evaluate $M.posBefore(e,L)$ and $M.posAfter(e,L)$ which return the index of the element e in the list of the specified type or the index of the first element $f$ of that type which appears before (or after) that element in the edit-string. This can be useful when we want to compute the first character that appears before a start marker for example. We will show that this can also be computed in time $O(\log(N))$.

More precisely, we will describe a new data type, which we call the Monotone Shared Edit Tree datatype which maintains 3 simultaneous representations of a user's edit-session corresponding to the standard, tracking changes, and edit-string views discussed above. This data type allows users to perform five different operations.
Two operations on the standard view (insert or delete a character) and three operations
on the edit-tree view (extend a node, insert a node into a node, or hide a character in a node). The operations are notated as shown below:
\begin{itemize}
\item $insert(k,\alpha)$ - insert character $\alpha$ at offset $k$ in the standard view and generate a corresponding edit-tree transformation $\tau$
\item $delete(k,\alpha)$ - delete the character $\alpha$ at offset $k$ in the standard view and generate a corresponding edit-tree transformation $\tau$
\item ${\tt treeextend}(v/m:p,\alpha)$ - extend the node $v/m:p$ by appending
a character $\alpha$ to the location $p$ at the end of the node and generate
the corresponding string transformation $\sigma$.
\item ${\tt treeinsert}(v/m:p,q,u/n,\alpha) $ -
 insert a new node $u/n$ containing the character $\alpha$
at position $q\le p$ in $v/m:p$ and generate
the corresponding string transformation $\sigma$.
\item ${\tt treehide}(v/m:p,q)$ - set
 the visibility of the character at position $q < p$
in  $v/m:p$ to false and generate
the corresponding string transformation $\sigma$. 
\end{itemize}
We will show that each of these operations can be computed in time $O(\log(N))$ where $N$ is the size of the edit string.

We will apply this to a collaborative editing system where each client
performs standard string edit-operations received locally from its user
and also performs tree-edit operations received from remote users.
For local edit operations, it converts them to edit-tree transformation, applies them 
to the local edit-tree and broadcasts the transformation to all peers. For remote edit-tree transformation, it applies them to its local tree, calculates the corresponding string operation and applies that locally to the standard view.

In the rest of this section we will formally define the MSET data type and prove that each of the local and remote operations can be performed in worst case time $O(\log(N))$

The MSET data type $M$ is implemented by maintaining the three list views using a three balanced binary trees where each in each tree node contains the number of elements of the specified type in its subtree. Looking up an element by its index and type can be performed in time $(O(\log(N))$.
Insertion of a new element between two elements of the list can be performed by adding a node to all three trees and rebalancing if necessary while maintaining the counts of number of leaves beneath each node. This can be done in $O(\log(N))$ time as well. In addition to maintaining these three lists, the system also maintains an edit tree representation and stores, for each element, the node and offset in which it appears in the tree. The key idea is to implement the edit tree in such a way that each local and remote editing operation can be performed in time $O(\log(N))$ while preserving the relationship between the three views and the edit tree.

We will represent each of the characters (visible, hidden, or marker) uniquely as an
element of type Atom.  With this approach, the lists $S$ and $T$ are sublists of $E$.

An Atom element will have the following fields:
\begin{itemize}
\item $nodepos$ this takes values of the form $(u/i/p)$ which specify the position $p$ of the atom in the $ith$ node created by user $u$.
\item $visible$ - this boolean value is true if the character is visible, false otherwise. It is false for the marker symbols and characters are "deleted" by setting this field to false.
\item $character$ - this is the character at that position which is either a character typed in by the user, or is a start marker $<^u_i$ or an end marker $>^u_i$.
\end{itemize}
Optionally, we can add a "node" field which gives a direct link to the corresponding node in the edit tree. This reduces look time from $O(\log(N))$ to $O(1)$ but
doesn't improve the asypmptotic performance so we will skip it in this version.

Each node $N$ in the edit tree will be represented by the following components:
\begin{itemize}
\item $N.user$ which is the user who created this node
\item $N.nodecount$ which is the number of nodes created by the user before this one
\item $N.atom$ which is an arraylist where $N.atom[i]$ will be a link to the $i$th atom in the node, where $i$ is a position consisting of a non-negative integer.
\item $N.start$ the atom holding the start marker for this node
\item $N.end$ the atom holding the end marker for this node
\item $N.insertionSet$ where $N.insertionSet[k]$ is a balanced binary tree map $S$ of the nodes inserted at position $k$ in the node.
Recall that nodes are inserted at each insertion point in a list ordered by their
userids. $S(u)$ gives the node $n$ owned by user $u$ and inserted at that position, if it exists; or gives a pair of nodes $n1,n2$ on either side of the position where a node owned by user $u$ should be inserted. We need this data structure to be able to rapidly process remote insertions with conflicts at the same position in a node. 
\end{itemize}
The insertion set is the key innovation which is needed to obtain logarithmic performance. It is required when processing an insertion operation at a conflict point, i.e. a point where several other users have simultaneously inserted an element. Since one could have $O(N)$ possible users, and these collision nodes are all ordered by user id, one needs to be able to lookup the correct position in time $O(\log(N))$ and hence some sort of balanced tree data structure is required.

We now present the MSET algorithm for each of the five possible edit operations: local insertion, local deletion, remote extension of a node, remote insertion of a node, and remote hiding of a character in a node.

\subsection{Insertion of a character in standard string view}

In this section we show how implement the local insert operation 
\[
insert(k,\alpha)
\]
for user $u$ assuming that it has already created $n$ nodes and that $\alpha$ has length $d=1$. This will get converted into one of the following two operations:
\begin{itemize}
\item ${\tt treeextend}(v/m:p,\alpha)$ 
\item ${\tt treeinsert}(v/m:p,q,u/n:1,\alpha) $
\end{itemize}
and will create a new element $e$ that gets inserted into each of the three lists
$M.S$, $M.R$, and $M.E$.

We must determine which type of transform is appropriate and to find the node $v/m:p$ and possibly the offset $q$ 
in time $O(\log(N))$.
Once these are calculated, they can be applied in time $O(\log(N))$ by using
a balanced binary tree search to find the node with id $v/m:p$ and the extension
of that node or the insertion of a new node can then be completed in time $O(d)$.

Lets look at the simple cases first, that is where we are inserting at the
beginning or the end of the string.

In the insertion is at the beginning ($k=0$), we need to find the first non-marker (hidden or visible) element $A=M.R[0]$ in the "tracking changes" view which will be at offset 0 in a node $v/m:p$. This takes time $O(\log(N))$.
From $A$ we can lookup the node id $v/m:p$ in constant time so the operation gets transformed into 
$tree_insert(v/m:p,0,u/n:d,\alpha)$. The new node $e$ corresponding to the character can then be created and inserted into $M.S$, $M.R$, and $M.E$ in
$O(\log(N))$ time.

This case of inserting at the end is similar, except that we need to find the last element $A$ in the tracking changes view.

This leaves the cases involving inserting into position $k$ in the middle of the string.
The first step is to look up the atoms $A_1=M.S[k]$ and $A_2=M.S[k+1]$ on either side of
the insertion point in the standard view and we then calculate the offset $k'$ of $A_1$ and $k''$ of $A_2$ of these atoms in the edit-string view and
the element $A' = M.E[k'+1]$ following $A_1$. This takes
$O(\log(N))$ time. There are three cases to consider.

>>>>

The data type $M$ allows the following operations:
\begin{itemize}
\item $M.treeextend(v/m:p,\alpha)$ - lookup the node with nodeid v/n:p and extend it by adding the character $\alpha$ at position $p$, update the lists $M.S$, $M.R$, and $M.E$ as well.
\item $M.treehide(v/m:p,q)$ - set the visibility to false for the element at position $q$ of the node with id $v/m:p$
\item $M.treeinsert(v/m:p,q,u/n,\alpha)$ - create a new node with id $u/n:1$ containing the single visible character $/alpha$ and insert that new node at offset $q$ in the node with id $v/m:p$.
\item $M.converttotreeop(\sigma)$- returns a tree-edit operation $\tau$ on $M.T$ that corresponds to the string edit operation $\sigma$ on $M.S$
\end{itemize}
We will provide algorithms for all five of these operations and prove that their complexity is $O(\log(N))$.
\subsection{implementation of characters and markers}
We assume all visible characters come from an alphabet $\Sigma$
and we expand that alphabet to $\Sigma^*$ by including all start and endmarkers.
The elements in the lists and the tree will be represented by a tuple that contains links to the nodes and also links to an indexing object $M.I$ which will allow one to rapidly calculate the position of the element in the lists (i.e. the number of elements of type $S$, $R$, or $E$ to the left of the element) and also allows us to rapidly find the $k$th element of each of the three lists.

@@

\begin{figure}[h]
\centering
\includegraphics[width=4.0in]{MSETfig002.jpg}
\caption{Index tree for the three lists. \label{fig:indextree}"}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=4.0in]{MSETfig001.jpg}
\caption{MSET tree and lists\label{fig:MSETtree}"}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=4.0in]{insertAtFront002.jpg}
\caption{Inserting a character at the beginning and end of $M.S$ \label{fig:frontinsert}"}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=4.0in]{insertMiddle001.jpg}
\caption{Cases 3a,3b: Inserting between two characters in $M.S$ \label{fig:midinsert}"}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=2.0in]{insertMiddle002.jpg}
\caption{Case 3c: inserting between two characters of $M.S$ \label{fig:frontinsert}"}
\end{figure}

@@

\subsection{applying $treeinsert(v/m:p,q,u/n,c)$ to $M.T$}
In this case, we need to lookup the target node $A = M.lookupNode(v/m:p)$ create a new node $B$ containing a single element
$e$ which contains the visible character $c$ and then we need to insert $B$
into the insertion set $S=A.iset[q]$ of the target node $A$ in time $O(\log(N))$ 
and to complete this operation we only need to insert the element $e$ into the correct place in $M.E$

\begin{figure}[h]
\centering
\includegraphics[width=1.0in]{insertTree001.jpg}
\caption{Inserting into $M.T$ when $S$ is empty \label{fig:emptyS}"}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=2.0in]{insertTree002.jpg}
\caption{Inserting into $M.T$ when $S$ is non-empty \label{fig:nonemptyS}"}
\end{figure}

Lets first consider the case where $S$ is empty, that is, nothing has yet been inserted. In this case $e$ will appear directly after the start marker $f = B.start$,
so we invoke $M.E.insertAfter(e,f)$.  Similarly, if $q=p$ so we are inserting at the end of node $B$, then $e$ appears directly before the end marker $f=B.end$,
so we invoke $M.E.insertBefore(e,f)$. Both operations can be performed in $O(\log(N))$ time.

If $S$ is not empty, then when we insert $B$ into it, it either appears as the first node in the insertion set, or the last node, or it appears between two nodes $C$ and $D$. In all of these cases, $B$ is either immediately before or immediately after another node $C$.

If $B$ is before a node $C$, then we insert $e$ before the start marker of $C$ with
$M.E.insertBefore(e,C.start)$. If $B$ appears immediately after a node $C$ in $S$,
then we insert $e$ after the end marker of $C$ with
$M.E.insertAfter(e,C.end)$.

\begin{theorem}
The implemenation of the MSET data type described above requires at most
$O(\log(N))$ time to process any of the five operations, that is,
the two standard string operations (insert and delete) and
the three edit-tree operations (treeextend, treeinsert, treehide), where $N$ is the size of the edit-tree before the operation is performed.
\end{theorem}

@@

In this section we give a detailed description of how to use the MSET data type to implement a convergent collaborative editing system among a group $U = \{u_1,\ldots,u_r\}$ of users.

In the simplest model, all users start off with an empty MSET. We assume that the users can send messages to all of the peers using a broadcast command.

Each user performs local edits on his or her string. These string edit operations are converted into tree edit operations and are applied to the users tree as well as being broadcast to all of the peers.  Users collect tree edit operations from their peers in a queue $Q$. When they process a tree edit operation $\tau$ the first check to see if its target node $A$ is in their tree $M.E$.  If not, then the operation is added to a set $M.I(A)$ of operations waiting for target node $A$.  If $A$ is in the tree, then $\tau$ is applied and it produces a result node $B$. The elements in $M.I(B)$ that were waiting for $B$ to be created are now added to the front of $Q$.

We call such a system the Basic MSET Collaborative Editing session. In the previous section we proved that each editing operation requires time at most $O(\log(N))$ to process by both the user orginating the edit operation and all of the peers processing it on their own trees.  Since insertion into a list requires time $O(\log(N))$ we see that, in terms of the amount of time to process each edit operations, this is an optimally efficient algorithm.

\section{Potential Pitfalls with MSET Collaborative Editing}

\subsection{Editing overload}
If too many users are editing a document at the same time, it can result in such a high load of edit operations that the user's will not be able to keep up.

For example, suppose that there are $R$ users each generating $G/\log(N)$ operations per second (where the logarithmic term accounts for the slowdown that occurs as the tree gets larger). Further, suppose that each user can process $P/\log(N)$ operations per second. Then, the rate at which edit operations are generated in the collaborative editing session is $R*G/\log(N)$ and users will
not be able to keep up if this exceeds their processing rate, i.e. if $P<R*G$.
Thus, if we assume that $P$ and $G$ are some standard rates for users on
typical hardware, this puts a limit on the number of users the system can support based on the average values of $P$ and $G$, viz. $R < P/G$.

This can be solved by using faster hardware (i.e. increase $P$) but is a fundamental
problem with scalability in collaborative editors.  There is also the cognitive overload problem of trying to imagine editing a document with 100 million simultaneous co-editors.

\subsection{Bottlenecks caused by a connection slowdown}
Another class of problem arises if one of the peer-to-peer connections, say from $u_1$ to $u_2$ becomes too slow. This can create a bottleneck for $u_2$ which
can effectively freeze all outside editing operations. 
There are however some potential problems. For example, suppose that all of the connections between the users are fast, except for the connection between $u_1$
and $u_2$ and suppose that $u_1$ performs 
an editing operation $\tau$ that creates a node $A$ and it sends that operation rapidly to
all users $u_3,\ldots,u_n$ but there is a very long delay before it can send $\tau$ to $u_2$.  In the meantime, assume that
all of the other users $u_3,\ldots,u_n$ are building upon
this target node. The user $u_2$ will receive and need to enqueue all of the operations of users $u_3,\ldots, u_n$ until it receives this "bottleneck" operation $\tau$. When $\tau$ finally is
received and node $A$ is created, all of the operations waiting on $A$ and its descendants must be processed.  It may take a substantial amount of time for $u_2$ to process the backlog
even though each individual operation requires only $O(\log(N))$ steps.

Such a situation can be minimized by allowing users to send "pull" requests to peers for edit operations that create a particular target, but this can increase the network traffic and slow down the system as well.

\section{Practical Implementations}
We have implemented a variant of the MSET data type and used it to create plugins for standard IDEs (jEdit, Eclipse, Netbeans) and our own editors \cite{granville_collabed:_2009}.  This variant introduced several
practical enhancements to improve performance (though not asympotically)
and to improve the user experience.  In this section, we briefly describe these
enhancements.

\subsection{Subnodes}
The first enhancement was to represent the lists $M.S$, $M.R$, and $M.T$ as lists of subnodes rather lists of single characters.  A subnode is a sequence of continguous characters in a node which is not interrupted by an non-null insertion set and in which all of the characters have the same visibility (i.e. the are either all visible or all hidden).  This introduces the extra complication of having to manage the list of subnodes with in a node (and to split a subnode if there is an insertion with in it or if one of the elements inside is set to hidden from visible).  The advantage of this approach is that copy and paste operations can be implemented as a single edit operation that creates a new node containing an entire string, not just a character.
For example, when a new file is opened, by the editor the tree will add a single node
containing one large string, and the user will send out one edit operation containing that string to all peers.

Likewise, deletion can be implemented by having a single visibility flag on an entire subnode so that cutting a large section of text can be done by setting the subnodes visibility flags to false rather than setting each individual elements flag to false.

\subsection{Attributes}
Building on the subnode model, we extended MSET to allow for each subnode to contain a set of attributes (e.g. font, style, color, etc.). To handle the inevitable conflicts, we used a protocol where each attribute had an associated version number and associated user id of the user who set that attribute's value.  When there is a conflict, the edit operation with the highest version number wins, and if there is a tie, then the one with the highest user id wins. If attribute version numbers are
represented as fixed size fields, then the number of times an attribute can be changed is limited.  If the version number is a "bignum" of unbounded size, then there is no maximum number of attributes but the space (and time) complexity of the system increases logarithmically in the version number.

\subsection{Manual Queueing of Incoming and/or Outgoing Edit Operations}
We have found it useful to maintain queue of incoming and outgoing edit operations
and to allow the user to temporarily stop processing incoming edit operations (and hence allowing the incoming queue to grow).  This is useful if the user want to perform some edits without being distracted by other users, for example, applying an emacs style repeated command that will process every line of a file with some particular sequence of edit operations.  Likewise, users may sometimes want to stop sending their edits to other users temporarily (and thereby allowing the outgoing edit queue to grow).  This can happen if they want to create a section of the document entirely by themselves and don't want any other user to see it until it is in a more finished form.  The MSET data type makes it very easy to implement these queues with no additional penalty beyond the time delay it will take to process the edit operations later rather than immediately. It has also turned out to be a feature that users often employ.

\section{Future work}
Two topics that we have not discussed are the implementation of undo operations
and the implementation of cut and paste operations.  Undo can be implemented fairly easily by performing a character insertion to undo a character deletion, and vice versa, but this does not revert the MSET to a previous state, rather it grows the state.  It might be possible to do better than this, at least in special cases.

Cut and paste operations are tricky because if another user performs a cut and paste on the code in which you are currently editing, the natural effect would be for you to continue editing the code but have it transferred to a different part of the document.  The current system, based on single character insertion and deletion operations, would simply steal all of the characters from around you and move them somewhere else. This is not very natural, but it seems to be quite difficult to implement a "natural" cut and paste algorithm for MSET Collaborative editors and this is an area that ought to be considered.
